Peaks Over Threshold for Bursty Time Series

### 2018-04-17 article_springer/article_springer.Rmd
"Statistics of the Mittag-Leffler Distribution"

output: rticles::springer_article

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = TRUE,
  message = FALSE,
  fig.height = 3, out.width = '\\textwidth'
  )
library(CTRM)
library(magrittr)
```

@@

# Scaling limit of Exceedance Times

@@

@@

Equation \eqref{eq:Tellscale} implies that for large $\ell$ we may 
approximate the distribution of $T(\ell)$ with an 
${\rm ML}(\beta, b(1/p))$ distribution.
Building on the POT (Peaks Over Threshold) method, we propose the 
following estimation procedure for the distribution of $T(\ell)$:

>>>>

output: 
  rticles::springer_article: 
    fig_caption: yes

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = TRUE,
  message = FALSE,
  fig.height = 3, out.width = '\\textwidth'
  )
library(CTRM)
library(MittagLeffleR)
library(magrittr)
```

@@

# Scaling limit of Exceedance Times {#sec:scaling}

@@

#### _Proof of Theorem:_
We interpret the threshold crossing time $T(\ell)$ as the hitting time of the 
underlying CTRM (Continuous Time Random Maxima) or "max-renewal process", 
and then utilize a result by @MeerschaertStoev08. 
The running maximum process is defined as 
$$
M(c) := J_1 \vee \ldots \vee J_{\lfloor c \rfloor},
$$
and since we assume that the $J_k$ have a continuous distribution, there exist 
norming functions $a(c)$ and $d(c)$ such that 
$$
\mathbf P\left[ \frac{M(c) - d(c)}{a(c)} \le \ell^* \right] 
\longrightarrow F(\ell^*), \quad t \to \infty
$$
where $F$ is a generalized extreme value distribution, and $\ell^*$ is any 
value from the support of $F$. 
The CTRM process is then defined via 
$$
V(t) = M(N(t)), \quad t \ge 0
$$
where $N(t)$ is the renewal process associated with the waiting times 
$W_k$: 
$$
N(t) = \max\{n: W_1 + \ldots + W_n \le t\}.
$$
Now a key observation is that
$$
T(\ell) = \inf\{t: V(t) > \ell\}, 
$$
and that 
$$
T(\ell) > t \quad \text{ if and only if } \quad V(t) \le \ell.
$$
By [Theorem 3.1, @MeerschaertStoev08], we have the stochastic process 
convergence 
$$
\frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\stackrel{d}{\longrightarrow} Y(t), \quad t > 0.
$$
where $Y(t)$ is a time-changed ("subordinated") extremal process, and where 
$\tilde b(c)$ is a regularly varying norming function which is
_inverse_ to $b(c)$, in the sense that 
$b(\tilde b(c)) \sim c \sim \tilde b(b(c))$.

Without loss of generality, we choose $\ell^*$ such that $F(\ell^*) = 1/e$, 
and let $\ell = a(\tilde b(c)) \ell^* + d(\tilde b(c))$. 
We may then calculate 
$$
\mathbf P\left[ \frac{T(\ell)}{b(1/p)} > t \right]
= \mathbf P[T(\ell) > b(1/p) t]
= \mathbf P[V(ct) \le \ell]
= \ldots
$$
where we have substituted $c = b(1/p)$, 
$$
\ldots
= \mathbf P\left[ \frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\le \frac{\ell - d(\tilde b(c))}{a(\tilde b(c))} \right]
\longrightarrow \mathbf P\left[ Y(t) \le \ell^* \right]
= \ldots
$$
Defining the hitting time of level $\ell^*$ by $Y(t)$ as 
$\xi_{\ell^*} := \inf\{t: Y(t) > \ell^*\}$,
we then have 
$$
\ldots = \mathbf P[\xi_{\ell^*} > t] 
= \mathbf P[(-\log F(\ell^*))^{-1/\beta} X^{1/\beta} D > t]
$$
by [Proposition 4.2, @MeerschaertStoev08], where $X$ is an exponential 
random variable with mean $1$. 
Using [Theorem 19.1, @Haubold11], we see that 
$X^{1/\beta} D \sim {\rm ML}(\beta, 1)$, concluding the proof. 
\qed

# Inference for the Mittag-Leffler distribution

```{r fig:QQ-est, fig.cap="QQ-estimator plots for various datasets.\\label{fig:QQ-est}", fig.height=7, fig.width=7}
par(mfrow = c(2,2))
flares_exponent <- flares %>% ctrm() %>% thin(k=500) %>% qqestplot(static=TRUE, main = "QQ-static, flare magnitudes")
ML_exponent <- MittagLeffleR::rml(500, 0.8) %>% qqestplot(static=TRUE, main = "QQ-static, Mittag-Leffler data")
ML_exponent2 <- MittagLeffleR::rml(500, 0.8) %>% qqestplot(static=FALSE, top_k = 15, conf.int = TRUE, main = "lala")
exp_exponent <- rexp(500) %>% qqestplot(static=TRUE, main = "QQ-static, Exponential data")
```

Since the Mittag-Leffler distribution is heavy-tailed, many researchers 
would intuitively give the highest importance to the tail behaviour of the 
distribution, and 
estimate the exponent of the tail function with established methods such 
as the Hill estimator. The QQ-estimator [@Kratz96] is closely related to 
the Hill estimator and fits a 
least squares line through the logarithms of the ordered statistics 
(y-axis) and the corresponding quantiles of the exponential distribution 
(x-axis), and returns the reciprocal slope as the estimate of the tail 
exponent.
For instance, the left panel in Figure \ref{fig:QQ-plots} shows the 
(static) QQ-estimator for the magnitudes of the solar flare data, returning 
a good fit to a Pareto distribution with tail parameter 
`r round(flares_exponent, 2)`.

However, as described nicely by @Resnick97, the less 
similar a heavy-tailed distribution is to a Pareto distribution, the less
useful a Hill estimator is. 
The top-right panel of \ref{fig:QQ-plots}, for instance, shows the 
QQ-estimator plot for $2000$ draws from the Mittag-Leffler distribution with 
tail parameter $0.8$. The stretched exponential shape means 
that the Mittag-Leffler distribution has more probability near $0$ than 
the Pareto distribution, severely biasing the estimator downwards 
(`r round(ML_exponent, 2)`). Hence a cutoff needs to be introduced, 
above which the tail exponent is fitted: The dynamic QQ-estimator 
plots the static QQ-estimates (y-axis) for different cutoffs (x-axis)
(see bottom-left panel). There is no apparent interval of stability, 
which could be taken to indicate a reasonable estimate for the true 
tail exponent $0.8$. 
Even worse, from a QQ-estimator point of view, there is no striking 
difference between the infinite mean ${\rm ML}(0.8,1)$ distribution and 
the light-tailed ${\rm Exp}(1)$ distribution (see bottom-right panel). 
In particular, if only few 
(e.g. $<100$) data are available, then the static QQ-estimator plots look 
virtually interchangeable.

Hence if there is some prior expectation that the data are drawn from the 
Mittag-Leffler distribution
(as is the the case for threshold exceedance times),
then we recommend avoiding QQ-estimators and Hill plots altogether, 
and instead examining QQ-plots directly 
(see Figure \ref{fig:QQ-plots}). 
The scale parameter $\sigma$ is irrelevant for a QQ-Plot. 
The tail parameter $\beta$ can be estimated quickly via the log-moment 
method by @Cahoy2013, or via maximum likelihood. 
Both estimators are implemented in the R package `MittagLeffleR` 
[@MittagLeffleR].

```{r QQ-plots, fig.cap="QQ Plots for Mittag-Leffler data (left; $\\beta = 0.8$) and threshold crossing times / magnitudes for the solar flare data (middle / right; threshold is set at the 500th order statistic, $\\beta = 0.83$). Threshold crossing times appear to fit the Mittag-Leffler distribution well, whereas the Magnitudes appear to fit a Pareto distribution better (\\ref{fig:QQ-est}). "}
par(mfrow = c(1, 3))
rml(n = 500, tail = 0.8) %>% mlqqplot(tail = 0.8, log = 'xy', main = "Mittag-Leffler data")
flares_arrivals <- flares %>% ctrm() %>% thin(k = 500) %>% interarrival()
flares_arrivals_MLparam <- (flares_arrivals %>% mlmle())$par
mlqqplot(flares_arrivals, 
         tail = flares_arrivals_MLparam["tail"], 
         scale = flares_arrivals_MLparam["scale"],
         log = 'xy', 
         main = "Threshold crossing times")
flares %>% ctrm() %>% thin(k = 500) %>% coredata() %>% 
  mlqqplot(tail = 0.8, log = 'xy', main = "Magnitudes")
```

@@

The Theorem in the [previous Section](#sec:scaling) implies that for high 
$\ell$ we may 
approximate the distribution of $T(\ell)$ with an 
${\rm ML}(\beta, b(1/p))$ distribution.
Building on the POT (Peaks Over Threshold) method, we propose the 
following estimation procedure for the distribution of $T(\ell)$:
